---
title: "PCA Analysis: Group F"
author: ""
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


In this document, we perform a Correspondence Analysis (CA) and Principal Component Analysis (PCA) on a specific group of animal species data labeled as **Group F**. The group consists of:
- bostaur
- oviscap+ovisari+caprhir (combined into 'ovi')
- susscrd
- cervela
- caprcap
- bosprim
- susscrf
- lepueur

```{r load-data}
library(FactoMineR)
library(sf)

alldata <- readRDS("tableTaxaPerMetasites.RDS")
grid_lim <- readRDS("grid_europe_lim.RDS")
grid <- readRDS("grid_europe_lim.RDS")

ovi <- apply(alldata[,c("oviscap","ovisari","caprhir")],1,sum)
alldata <- cbind(alldata, ovi)

groupf <- alldata[,c("bostaur","susscrd","cervela","caprcap","bosprim","susscrf","lepueur","ovi")]

Namef <- paste0(c("oviscap-ovisari-caprhir", "bostaur", "susscrd", "cervela", "caprcap", "bosprim", "susscrd", "susscrf", "ovi"), collapse="//")
#empty <- which(apply(groupf, 1, sum) == 0 & apply(groupf, 1, sum) < 100 )
empty <- which(apply(groupf, 1, sum) == 0 )#& apply(groupf, 1, sum) < 100 )
```

## Some Basic Analysis

```{r}
inter  <-  st_intersects(st_make_valid(grid),allsite_sf)
grid_lim <- grid[lengths(inter)>0] 
inter  <-  inter[lengths(inter)>0]

##simple plot to show the number of sites per grid
plot(st_bind_cols(grid_lim,len=log(lengths(inter)))[,"len"],main=paste("(log) #Site per Hexagons"))
plot(st_bind_cols(grid_lim,len=log(apply(alldata,1,function(i)sum(i>0)))),main="(log) #num taxa per Hex")
plot(st_bind_cols(grid_lim,len=log(apply(alldata,1,sum)))[,"len"],main="(log) total #nisp per Hex")
```




## Correspondence Analysis

### Initial CA Plot

```{r ca-initial}
caGroupeF <- CA(groupf[-empty, ])
plot(st_bind_cols(grid[-empty, ], caGroupeF$row$coord[, 1]), reset = FALSE, main = paste("CA1", Namef))
plot(wd, add = TRUE, border = "red", lwd = 2)
```

### CA Minus Outliers

```{r ca-minus-outliers}
empty <- c(empty, which(apply(groupf, 1, function(i) any(i > apply(groupf[-empty, ], 2, quantile, probs = 0.95)))))
caGroupeF <- CA(groupf[-empty,])
ca1 <- caGroupeF$row$coord[, 1]
ca1 <- ca1 - min(ca1)  + 1
plot(st_bind_cols(grid[-empty, ], log(ca1)), reset = FALSE, main = "CA1 minus outliers")
plot(wd, add = TRUE, border = "red", lwd = 2)
```

Using the percentage

```{r}
percentage <- groupf[-empty,]
tot <- apply(percentage,1,sum)
percentage <- apply(percentage,2,function(i)i/tot)
pca.perc <- FactoMineR::PCA(log(percentage+1))
plot(st_bind_cols(grid[-empty, ], pca.perc$ind$coord), reset = FALSE, main = "CA1 minus outliers")
plot(wd, add = TRUE, border = "red", lwd = 2)

alltaxon  <-  query.database(conn = conn,sql.command = "SELECT * FROM zoptions_TaxaList")

onlyMammal  <-  alldata[colnames(alldata) %in% alltax
```


